---
title: "Hands_on_HOMEWORK_III"
author: "Marko Ludaic"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Exercise 1

[Stamey et al. 1989](https://www.auajournals.org/doi/10.1016/S0022-5347%2817%2941175-X) examined the correlation between the level of prostate-specific antigen (PSA) and a number of clinical measures in men who were about to receive a radical prostatectomy. PSA is a protein that is produced by the prostate gland. The higher a man’s PSA level, the more likely it is that he has prostate cancer.  
Use the [prostate cancer dataset](data/prostate_data.txt), described [here](data/prostate_description.txt),  to train a model that predicts log of prostate-specific antigen. 
The variables are    

- log cancer volume (lcavol)  
- log prostate weight (lweight)  
- age  
- log of the amount of benign prostatic hyperplasia (lbph)   
- seminal vesicle invasion (svi)  
- log of capsular penetration (lcp)  
- Gleason score (gleason)    
- percent of Gleason scores 4 or 5 (pgg45)  

You can ignore column named "train" and do your own data splitting.  
Do not forget to perform feature selection!   
You can use as examples the [Linear Regression Lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch3-linreg-lab.html) and the section related to feature selection from  [Lab: Linear Models and Regularization Methods
](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch6-varselect-lab.html) from the book [An Introduction to Statistical Learning](https://www.statlearning.com/).

```{r}
# Loading packages
library(leaps)
library(caTools)
```


```{r}
# Load data and ignore train column
prostate_data <- read.csv("data/prostate_data.txt", sep = "")
prostate_data <- subset(prostate_data, select = -train)

is.na(prostate_data) # No missing value

# Split the data into training and test sets
set.seed(123) 
split <- sample.split(prostate_data$lpsa, SplitRatio = 0.8)
training_set <- subset(prostate_data, split == TRUE)
test_set <- subset(prostate_data, split == FALSE)
```

- We select 80% of the data to be used for training the model and the remaining 20% for testing it.
Now we try to predict by doing a linear regression model with all our features:

```{r}
# Fit the linear regression model with all the predictors
regressor <- lm(lpsa ~ ., data = training_set)

# Make predictions on the test set
y_pred <- predict(regressor, newdata = test_set)

y_pred

# Quantify accuracy of the model
rmse <- sqrt(mean((y_pred - test_set$lpsa)^2))
rmse # 0.6432299
```
- We have to take into account that we need to do feature selection first, to reduce overfitting and improve our model. 
Our criteria will be having the maximum R-squared value (or Adjusted in our case), and the minimun BIC statistic, so we are going to determine how many features accomplish that:

```{r}
regfit.full <- regsubsets(lpsa ~ ., data = training_set)
reg.summary <- summary(regfit.full)
reg.summary$rsq # r-squared depending on number of features

par(mfrow = c(1, 2))
plot(reg.summary$adjr2, xlab = "Number of features",
     ylab = "Adjusted RSq", type = "l")

which.max(reg.summary$adjr2) # Maximum R-squared with 6 features

plot(reg.summary$cp, xlab = "Number of features",
    ylab = "Cp", type = "l")

which.min(reg.summary$cp) # Minimum Cp  with 3 features

plot(reg.summary$bic, xlab = "Number of features",
    ylab = "BIC", type = "l")

which.min(reg.summary$bic) # Minimum BIC statistic with 3 features


coef(regfit.full, 3) # See coefficients and features selected
coef(regfit.full, 6)
```
- We have two options, 3 or 6 features, so we will try both for building the linear regression model and see their accuracy on the predictions:

```{r}
# 3 features
regressor3f <- lm(lpsa ~ lcavol + lweight + svi , data = training_set)

y_pred_3f <- predict(regressor3f, newdata = test_set)

y_pred_3f

rmse <- sqrt(mean((y_pred_3f - test_set$lpsa)^2))
rmse # 0.6779363

# 6 features
regressor6f <- lm(lpsa ~ lcavol + lweight + age + lbph + svi + pgg45, data = training_set)

y_pred_6f <- predict(regressor6f, newdata = test_set)

y_pred_6f

rmse <- sqrt(mean((y_pred_6f - test_set$lpsa)^2))
rmse # 0.6503116

```
- Any of these number of features seemed to be better than our first attempt using all the variables, but in terms of RMSE this is not the case, so it could be that it is overfitting to the training data, and maybe the feature selection could not capture the complexity of the model (there could be some nonlinear relationships with some features, or that some important features were not included).

In any case, we propose the linear model with 6 features to avoid overfitting, as it the one with highest adjusted R-squared, so, based on the coefficients, we have:

```{r}
coef(regfit.full, 6)
```
- Here we have the final model regression:

lpsa =  0.553796794 + 0.508246857(lcavol) + 0.623766684(lweight) - 0.019102557(age) + 0.081727054(lbph) + 0.591146702(svi) + 0.003598437(pgg45)

----------------------------------------------------------------------------------------

# Exercise 2

Use the [breast cancer dataset](data/breat_cancer_data.csv) to train a model that predicts whether a future tumor image (with unknown diagnosis) is a benign or malignant tumor. Try different machine learning algorithms such as:   
- KNNs  
- Decision trees  
- Random forest  
- Logistic Regression  

The breast cancer dataset contains digitized breast cancer image features, and was created by [Dr. William H. Wolberg, W. Nick Street, and Olvi L. Mangasarian](https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Diagnostic%29). Each row in the data set represents an image of a tumor sample, including the diagnosis (benign or malignant) and several other measurements (nucleus texture, perimeter, area, and more). Diagnosis for each image was conducted by physicians.

Do not forget to perform hyperparameter tuning!   
Which of all models performs better for this data? Discuss.  

Generate a ROC curve for all the models. 

You can use as a guide the analysis of this dataset included in the [chapter 5](https://datasciencebook.ca/classification1.html) of the Data Science, A First Introduction Book.
Additionally, for further information and ideas, you can check [this post](https://www.rebeccabarter.com/blog/2020-03-25_machine_learning/)

```{r}
# Loading packages
library(tidyverse)
library(tidymodels)
library(yardstick)
library(ROCR)
library(rpart.plot)
library(ranger)
library(glmnet)
```

```{r}
# Load the dataset and take a quick view
breast_cancer <- read.csv("data/breat_cancer_data.csv")
glimpse(breast_cancer)

# Eliminate redundant id column and X column full of NA
breast_cancer <- subset(breast_cancer, select = -c(id,X))

# Turn the diagnosis into factor, and verify the categories there
breast_cancer <- breast_cancer |>
  mutate(diagnosis = as_factor(diagnosis))
glimpse(breast_cancer)

breast_cancer |>
  pull(diagnosis) |>
  levels() # "M" "B"

# Percentage of diagnosis types
num_obs <- nrow(breast_cancer)
breast_cancer |>
  group_by(diagnosis) |>
  summarize(
    count = n(),
    percentage = n() / num_obs * 100
  ) # 37% M and 63% B
```

- We take a quick look to the data and our response variable.

```{r}
# Split the data into training and testing sets
set.seed(123)

bc_split <- initial_split(breast_cancer, prop = 0.70, strata = diagnosis)
bc_train <- training(bc_split)
bc_test <- testing(bc_split)

# Check proportion of diagnosis in each set
bc_train |>
  group_by(diagnosis) |>
  summarize(
    count = n(),
    percentage = n() / nrow(bc_train) * 100
  ) # 37.3% M and 62.7% B

bc_test |>
  group_by(diagnosis) |>
  summarize(
    count = n(),
    percentage = n() / nrow(bc_test) * 100
  ) # 37.2% M and 62.8% B
```

- 70% of the data goes for training and 30% for testing. We can observe that proportions for diagnosis are practically maintained in both sets.

```{r}
# Set up the recipe to preprocess the data
bc_recipe <- recipe(diagnosis ~ ., data = bc_train) %>%
  # downsampling to balance number of B and M
  themis::step_downsample(diagnosis) %>% 
  #data normalization
  step_center(-diagnosis) %>% 
  step_scale(-diagnosis) %>% 
  step_BoxCox(-diagnosis) %>% 
  modify_if(is.character, as.factor)%>%
  #apply the recipe to the data
  prep()

#apply the recipe to the training set
bc_juiced <- juice(bc_recipe)

#apply the same recipe to the test data
baked_test <- bake(bc_recipe, new_data = bc_test)
```

- Data is preprocessed and we obtain two new sets, bc_juiced for training and baked_test for test.

- Now it is time to generate models to predict the diagnosis:

# KNN 
```{r}
knn_spec <- nearest_neighbor() %>% 
  set_engine("kknn") %>% 
  set_mode("classification")

#Fit the pre-processed training data
knn_fit <- knn_spec %>% 
  fit(diagnosis ~., data = bc_juiced)

knn_fit

#First evaluation of the model
knn_fit %>% 
  predict(baked_test) %>% 
  bind_cols(baked_test) %>% 
  metrics(truth = diagnosis, estimate = .pred_class)
# 94.2% accuracy and 87.5% Kappa
```

- Accuracy and kappa metrics are used to evaluate this first model with default parameters. We proceed with hyperparameter tuning. In this case, the number of neighbors.

```{r}
# Hyperparameter tuning: Number of neighbors
#Create 5 folds for cross validation on the training data set
cv <- vfold_cv(bc_juiced, v = 5)

knn_tune <- nearest_neighbor(neighbors = tune()) %>% 
  set_engine('kknn') %>% 
  set_mode('classification')

knn_wf <- workflow() %>% 
  add_model(knn_tune) %>% 
  add_recipe(bc_recipe)

#create a grid of hyperparameter values to test
knn_grid <- tibble(neighbors = c(5, 10, 15, 20, 25, 30, 50, 75, 100, 125, 150))

#tune  workflow
set.seed(123)
knn_tuning <- knn_wf %>% 
  tune_grid(resamples = cv,
            grid = knn_grid,
            metrics = metric_set(accuracy, roc_auc,  kap))

#show the top 5 best models based on roc_auc metric
knn_tuning %>% show_best('roc_auc')
```

- We choose 20 neighbors as it is the model with higher AUC.

```{r}
knn_spec_20 <- nearest_neighbor(neighbors = 20) %>% 
  set_engine("kknn") %>% 
  set_mode("classification")

# Fit the pre-processed training data
knn_fit_20 <- knn_spec_20 %>% 
  fit(diagnosis ~ ., data = bc_juiced)

knn_fit_20

# Evaluation of the final model
knn_fit_20 %>% 
  predict(baked_test) %>% 
  bind_cols(baked_test) %>% 
  metrics(truth = diagnosis, estimate = .pred_class)
# 95.9% accuracy and 91.3% Kappa
```

- AUC plot

```{r}
# Generate predicted probabilities and predicted class labels for the test set
test_preds <- knn_fit_20 %>% 
  predict(baked_test) %>% 
  bind_cols(bc_test) %>% 
  select(diagnosis, .pred_class)

# Convert the factor variables to numeric vectors
test_scores <- as.numeric(test_preds$.pred_class == "M",test_preds$.pred_class =="B")
test_labels <- as.numeric(test_preds$diagnosis == "M",test_preds$diagnosis == "B")

# Generate the ROC curve using the ROCR package
test_pred <- prediction(test_scores, test_labels)
perf <- performance(test_pred, "tpr", "fpr", ret=c("fpr", "tpr", "cutoff"))

# Plot the ROC curve
plot(perf, main = "ROC Curve", col = "blue", lwd = 2, legacy.axes = TRUE)
```


# Decision trees  

```{r}
#Specify the model
dt_mod <- decision_tree() %>% 
  set_engine("rpart") %>% 
  set_mode("classification")

#create 10-fold cross-validation data object
cv <- vfold_cv(bc_train)

#Fit the decision tree model with 10-fold cross-validation and save all the model object
dt_default <- fit_resamples(dt_mod, 
                            preprocessor = bc_recipe, 
                            resamples = cv,
                            control = control_resamples(extract = function(x) extract_model(x))
)

#pulling out the model
pull_model <- function(extract_fold) {
  extract_fold$.extracts[[1]]
}

all_mods <- map(dt_default$.extracts, pull_model)

#Estimate the model's performance
collect_metrics(dt_default) # 91.9% accuracy and 93.1% AUC

# See some of the models generated
rpart.plot(all_mods[[1]], type = 4)
rpart.plot(all_mods[[2]], type = 4)
rpart.plot(all_mods[[3]], type = 4)
```

- The accuracy and the AUC is calculated for the evaluation of this initial model. 

```{r}
### Hyperparameter tuning ####
#First modify decision tree model
dt_tune <- dt_mod %>% 
  set_args(cost_complexity = tune(),
           min_n = tune())

#Use a regular grid with 10 possible values for cost complexity and 5 possible values for the minimum n.
dt_grid <- grid_regular(
  cost_complexity(), 
  min_n(), 
  levels = c(10, 5)
)

# Grid search
dt_tune_fit <- tune_grid(
  dt_tune,
  preprocessor = bc_recipe,
  resamples = cv,
  grid = dt_grid
) # It can take some time

#Showing the best hyperparameter combinations based on roc_auc metric
show_best(dt_tune_fit, metric = "roc_auc")
```

- Wee see that the best models have the same value for the roc AUC metric and min_n, so we choose the one with less cost complexity.

```{r}
#tuned model
dt_tune2 <- dt_tune %>% 
  set_args(cost_complexity = 0.0000000001, min_n = 21)

#fit the tuned model
dt_tune_fit2 <- tune_grid(
  dt_tune2,
  preprocessor = bc_recipe,
  resamples = cv,
)

best_params <- select_best(dt_tune_fit2, metric = "roc_auc")
final_mod <- finalize_model(dt_tune2, best_params)
final_mod

dt_finalized <- last_fit(final_mod,
                         preprocessor = bc_recipe,
                         split = bc_split)

dt_finalized$.metrics[[1]] # Now 92.4% accuracy and 95.7% AUC

```

- As expected, the evaluation shows now a better model. Now we plot again the ROC curve.

```{r}
# Extract the underlying model object
dt_model <- extract_fit_parsnip(dt_finalized)

# Generate predicted probabilities and predicted class labels for the test set
test_preds <- predict(dt_model, baked_test) %>% 
  bind_cols(bc_test) %>% 
  select(diagnosis, .pred_class)

# Convert the factor variables to numeric vectors
test_scores <- as.numeric(test_preds$.pred_class == "M")
test_labels <- as.numeric(test_preds$diagnosis == "M")

# Generate the ROC curve using the ROCR package
test_pred <- prediction(test_scores, test_labels)
perf <- performance(test_pred, "tpr", "fpr", ret=c("fpr", "tpr", "cutoff"))

# Plot the ROC curve
plot(perf, main = "ROC Curve", col = "blue", lwd = 2, legacy.axes = TRUE)

```


# Random forest  

```{r}
#Specify the model
rf_spec <- rand_forest() %>% 
  set_engine("ranger") %>% 
  set_mode("classification")

#Fit the pre-processed training data
rf_fit <- rf_spec %>% 
  fit(diagnosis ~., data = bc_juiced)

rf_fit

#Evaluate random forest model
rf_fit %>% 
  predict(baked_test) %>% 
  bind_cols(baked_test) %>% 
  metrics(truth = diagnosis, estimate = .pred_class)
# 92.4% accuracy and 84.17% kappa
```

- First evaluation seems to be not as good as with the previous ones.. Let's do hyperparameter tuning.

```{r}
#Hyperparameter tuning
rf_model <- 
  rand_forest() %>%
  set_args(mtry = tune()) %>%
  # select the engine/package that underlies the model
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification") 

#define 10-cross-fold validation on the training set
cv <- vfold_cv(bc_train)

# set the workflow
rf_workflow <- workflow() %>%
  # add the recipe
  add_recipe(bc_recipe) %>%
  # add the model
  add_model(rf_model)

#specify which values to try with
rf_grid <- expand.grid(mtry = c(3, 4, 5))

#Perform grid search with mtry values 3,4 and 5
rf_tune_results <- rf_workflow %>%
  tune_grid(resamples = cv,
            grid = rf_grid, # grid of values to try
            metrics = metric_set(accuracy, roc_auc,  kap) # metrics we care about
  )

show_best(rf_tune_results, metric = "roc_auc")
```

- The best mtry seems to be 4 based on the roc AUC, so we will use that model.

```{r}
#tuned model
rf_tuned <- 
  rand_forest() %>%
  # specify that the `mtry` parameter needs to be tuned
  set_args(mtry = 4) %>%
  # select the engine/package that underlies the model
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification") 

rf_tuned

# set the tuned workflow
rf_workflow_tuned <- workflow() %>%
  add_recipe(bc_recipe) %>%
  add_model(rf_tuned)

rf_finalized <- last_fit(rf_workflow_tuned,
                         preprocessor = bc_recipe,
                         split = bc_split,
                         metrics = metric_set(accuracy, roc_auc,  kap))

#evaluate tuned model
rf_tune_results2 <- rf_workflow_tuned %>%
  tune_grid(resamples = cv,
            metrics = metric_set(accuracy, roc_auc,  kap) # metrics we care about
  )

rf_tune_results2 %>%
  collect_metrics() # 94.9% accuracy, 89.5% kappa and 98.9% AUC
```

```{r}
# Plot AUC

rf_model <- extract_fit_parsnip(rf_finalized)

test_preds <- predict(rf_model, baked_test) %>% 
  bind_cols(bc_test) %>% 
  select(diagnosis, .pred_class)

# Convert the factor variables to numeric vectors
test_scores <- as.numeric(test_preds$.pred_class == "M")
test_labels <- as.numeric(test_preds$diagnosis == "M")

# Generate the ROC curve using the ROCR package
test_pred <- prediction(test_scores, test_labels)
perf <- performance(test_pred, "tpr", "fpr")


# Plot the ROC curve
plot(perf, main = "ROC Curve", col = "blue", lwd = 2, legacy.axes = TRUE)
```


# Logistic regression 

```{r}
#Defining a model
logistic_model <- 
  logistic_reg() %>%
  set_engine("glm", family = "binomial") %>%
  set_mode("classification")


# Fit the model
lr_fit <- logistic_model %>% 
  fit(diagnosis ~., data = bc_juiced)

tidy(lr_fit)

# Categorical predictor variables
x <- model.matrix(diagnosis ~., data = bc_juiced)[,-1]

# Convert the outcome (class) to a numerical variable
y <- ifelse(bc_juiced$diagnosis == "M", 1, 0)
```

```{r}
# Hyperparameter tuning (lambda)
cv_lasso <- cv.glmnet(x, y, alpha = 1, family = "binomial")
model <- glmnet(x, y, family = "binomial", alpha = 1, lambda = cv_lasso$lambda.min)
cv_lasso$lambda.min # Optimal number of lambda = 0.0087
# Display regression coefficients
coef(model)

# Make predictions on the test data
x_test <- model.matrix(diagnosis ~., baked_test)[,-1]
probabilities <- model %>% predict(newx = x_test)
predicted <- ifelse(probabilities > 0.5, "M", "B")

# Model accuracy
observed <- baked_test$diagnosis
mean(predicted == observed) # Accuracy of 97%
```

- We see a really good accuracy here in this model. Let's plot the ROC curve

```{r}
# Create a prediction object
predictions <- prediction(probabilities, baked_test$diagnosis)

# Calculate the ROC curve and AUC
roc <- performance(predictions, "tpr", "fpr")
auc <- performance(predictions, "auc")
auc <- auc@y.values[[1]]
auc # 99.8%

# Plot the ROC curve
plot(roc, main = "ROC Curve", col = "blue", lwd = 2, ylim = c(0, 1), xlim = c(0, 1))
abline(a = 0, b = 1, lwd = 2, lty = 2, col = "gray")
```

- After generating and evaluating all these models, we can conclude that the logistic regression model seems to be the best one in terms of AUC (99.81%) and accuracy (97%). Also the curve seems to be smoother compared to the other models.

# Exercise 3  

Use [The Cancer Genome Atlas (TCGA)](https://www.genome.gov/Funded-Programs-Projects/Cancer-Genome-Atlas) gene expression data of two different cancer types to build a machine learning model that identifies whether one unknown sample belongs to one or the other. The TCGA is a comprehensive and coordinated effort to accelerate our understanding of the molecular basis of cancer through the application of genome analysis technologies, including large-scale genome sequencing. The program has generated, analyzed, and made available genomic sequence, expression, methylation, and copy number variation data on over 11,000 individuals who represent over 30 different types of cancer. 
After building your model, you should predict the cancer types for [10 unkwnon samples](data/unknwown_samples.tsv).  

For this task, you should retrieve the TCGA data from the [Genomic Data Commons Data Portal](https://portal.gdc.cancer.gov/). If necessary you can watch the video uploaded in the Campus Global. The video assumes that you have previously installed the [GDC data transfer tool](https://gdc.cancer.gov/access-data/gdc-data-transfer-tool). 

Each team will work with two specific cancer types, that will be assigned in class.

Important notice: if you do not have a lot of hard drive space in your laptop, you can modify the manifest file to download only 50 samples per cancer types. 
As part of the assignment, you should provide the input data fed to the machine learning algorithm as a tsv file. 

```{r}
# Loading the packages
library(caret)
library(glmnet)
library(pROC)
library(randomForest)
```


```{r}
# loading the data
files <- list.files(path = "data/LAML/", pattern = "\\.tsv$")

for (i in seq_along(files)) {
  # Read the file into a data frame
  data <- read.table(paste("data/LAML/", files[i], sep=""), header=TRUE, sep="\t")
  
  # Delete rows 1, 3, 4, 5, and 6
  rows_to_delete <- c(1, 3, 4, 5, 6)
  data <- data[-rows_to_delete, ]
  
  # Open the remaining rows using read.table() and separating by \t
  assign(paste0("laml", i), read.table(textConnection(paste(apply(data, 1, paste, collapse = "\t"), collapse = "\n")), header=TRUE, sep="\t"))
}

##############################
# Loop over all the data frames
for (i in 1:50) {
  # Read the file into a data frame
  data <- read.table(paste("data/LAML/", files[i], sep=""), header=TRUE, sep="\t")
  
  # Assign the first column values as row names
  rownames(data) <- data[, 1]
  
  # Delete the first three columns
  data <- data[, -c(1:3)]
  
  # Assign the modified data frame to a variable with a name like laml1, laml2, etc.
  assign(paste0("laml", i), data)
}

###################

# Loop over all the data frames
for (i in 1:50) {
  # Read the file into a data frame
  data <- read.table(paste("data/LAML/", files[i], sep=""), header=TRUE, sep="\t")
  
  # Assign the first column values as row names
  rownames(data) <- data[, 1]
  
  # Delete the first three columns
  data <- data[, -c(1:3)]
  
  # Delete rows 1 to 4
  data <- data[-c(1:4), ]
  
  # Assign the modified data frame to a variable with a name like laml1, laml2, etc.
  assign(paste0("laml", i), data)
}


############ mean
# Load all data frames into a list
list_of_data_frames <- lapply(paste0("laml", 1:50), function(x) get(x))

# Calculate the mean for each slot across all data frames
laml_data <- Reduce("+", list_of_data_frames) / length(list_of_data_frames)


####################################################################### LGG ##################################################################### 
###################################################################### 


files <- list.files(path = "data/LGG/", pattern = "\\.tsv$")

for (i in seq_along(files)) {
  # Read the file into a data frame
  data <- read.table(paste("data/LGG/", files[i], sep=""), header=TRUE, sep="\t")
  
  # Delete rows 1, 3, 4, 5, and 6
  rows_to_delete <- c(1, 3, 4, 5, 6)
  data <- data[-rows_to_delete, ]
  
  # Open the remaining rows using read.table() and separating by \t
  assign(paste0("lgg", i), read.table(textConnection(paste(apply(data, 1, paste, collapse = "\t"), collapse = "\n")), header=TRUE, sep="\t"))
}

##############################
# Loop over all the data frames
for (i in 1:50) {
  # Read the file into a data frame
  data <- read.table(paste("data/LGG/", files[i], sep=""), header=TRUE, sep="\t")
  
  # Assign the first column values as row names
  rownames(data) <- data[, 1]
  
  # Delete the first three columns
  data <- data[, -c(1:3)]
  
  # Assign the modified data frame to a variable with a name like lgg1, lgg2, etc.
  assign(paste0("lgg", i), data)
}

###################

# Loop over all the data frames
for (i in 1:50) {
  # Read the file into a data frame
  data <- read.table(paste("data/LGG/", files[i], sep=""), header=TRUE, sep="\t")
  
  # Assign the first column values as row names
  rownames(data) <- data[, 1]
  
  # Delete the first three columns
  data <- data[, -c(1:3)]
  
  # Delete rows 1 to 4
  data <- data[-c(1:4), ]
  
  # Assign the modified data frame to a variable with a name like lgg1, lgg2, etc.
  assign(paste0("lgg", i), data)
}


############ mean
# Load all data frames into a list
list_of_data_frames <- lapply(paste0("lgg", 1:50), function(x) get(x))

# Calculate the mean for each slot across all data frames
lgg_data <- Reduce("+", list_of_data_frames) / length(list_of_data_frames)
```

- In this first part, we've prepared the TCGA data of both cancer types. We've used 50 samples of both cancer types, since the usage of more requires more computational power and storage. Further, we've deleted 5 rows that were not out of use for the analysis and training the model and assigned the correct rownames as genes. The same was done for both cancer types. 

```{r}
################ MERGING TWO DATA FRAMES
rownames(laml_data) <- paste0(rownames(laml_data), "_laml")
laml_data$cancer_type<-"0"
rownames(lgg_data) <- paste0(rownames(lgg_data), "_lgg")
lgg_data$cancer_type<-"1"
###
data<-rbind(laml_data,lgg_data)
###

data$cancer_type <- factor(data$cancer_type)
################################################################
# Center the data
data_centered <- scale(data[,1:6], center = TRUE, scale = FALSE)

# Scale the data
data_scaled <- scale(data_centered[,1:6], center = FALSE, scale = TRUE)
data_scaled<-as.data.frame(data_scaled)
data_scaled$cancer_type<-data$cancer_type
################################################################

###
set.seed(123)  # for reproducibility
train_index <- createDataPartition(data_scaled$cancer_type, p = 0.8, list = FALSE, times = 1)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]
###
ggplot(train_data, aes(x = cancer_type)) +
  geom_bar(fill = "#fc9272") +
  ggtitle("Distribution of cancer types in train_data") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(test_data, aes(x = cancer_type)) +
  geom_bar(fill = "#fc9272") +
  ggtitle("Distribution of cancer types in test_data") +
  theme_minimal() +
  theme(legend.position = "none")

cc<-t(train_data)
###
lm <- glm(formula = cancer_type ~ ., data = train_data, family = binomial())
summary(lm)
```

- In the next part, we've merged two data frames, so that we could get one data frame with the data containing both cancer types. As columns, we perserved same variables, but then added a new one which would indicate what cancer type the sample gene refers to. The column we added is "cancer_type" and it has the value of 0 for LAML and the vaule of 1 for LGG. The data was centered and scaled before being divided into train and test data. 

```{r}
### FILTERING
# Lasso Regression
# Split data into training and test sets
set.seed(123)
train_index <- createDataPartition(data_scaled$cancer_type, p = 0.8, list = FALSE, times = 1)
train_data <- data_scaled[train_index, ]
test_data <- data_scaled[-train_index, ]
# Convert the cancer_type column to a factor variable
train_data$cancer_type <- as.factor(train_data$cancer_type)
test_data$cancer_type <- as.factor(test_data$cancer_type)
```

```{r}
# Fit a Lasso model using glmnet
x <- model.matrix(cancer_type ~ ., data = train_data)[,-1] # exclude the intercept column
y <- train_data$cancer_type
fit <- cv.glmnet(x, y, family = "binomial", alpha = 1, nfolds = 5) # takes some time 
# Plot the cross-validated deviance as a function of lambda
plot(fit)
# Select the optimal lambda value based on cross-validation
lambda_opt <- fit$lambda.min
# Get the coefficients of the Lasso model at the optimal lambda
coef_opt <- coef(fit, s = lambda_opt)
coef_opt[coef_opt != 0]
```

- Here we've decided to use Lasso regression which is a type of linear regression that performs variable selection and regularization. It is commonly used in machine learning for feature selection, which can help improve the performance of a model and reduce overfitting. In our example, Lasso regression is used to build a predictive model for cancer type classification. The "glmnet" package is used to perform the Lasso regression, which is followed by splitting the data into training and test sets. The training set is used to fit the model, while the test set is used to evaluate the model's performance on unseen data.

```{r}
# Evaluate the performance of the Lasso model on the test set
x_test <- model.matrix(cancer_type ~ ., data = test_data)[,-1]
y_test <- test_data$cancer_type
pred <- predict(fit, s = lambda_opt, newx = x_test, type = "response")
auc <- roc(y_test, pred)$auc
auc # 0.9696
```

- We have obtained the AUC value of ~ 0.97 which we considered a good and reliable.

```{r}
## Prepare unknown samples

# Load and clean unknown data
unknown_data <- read.table("data/unknown_samples.tsv", header = TRUE, row.names = 1, sep = "\t")
unknown_data <- unknown_data[, -1]

# Preprocess unknown data
# Center the data
unknown_data_centered <- scale(unknown_data[,1:10], center = TRUE, scale = FALSE)

# Scale the data
unknown_data_scaled <- scale(unknown_data_centered[,1:10], center = FALSE, scale = TRUE)

```

```{r}
# Load the data for the individual to predict on
individual_1 <- data.frame(unknown_data_scaled[,1])
colnames(individual_1) <- "tpm_unstranded"

input_data_1 <- matrix(0, nrow = nrow(individual_1), ncol = 6)
input_data_1[,4] <- individual_1$tpm_unstranded
rownames(input_data_1) <- rownames(individual_1)
input_data_1 <- data.frame(input_data_1)
colnames(input_data_1) <- c("unstranded", "stranded_first", "stranded_second", "tpm_unstranded",
                            "fpkm_unstranded", "fpkm_uq_unstranded")
input_matrix_1 <- as.matrix(input_data_1)

# Make predictions using the model
predictions_1 <- predict(fit, s = lambda_opt, newx = input_matrix_1, type = "response")
predicted_1 <- ifelse(predictions_1 > 0.5, 1, 0)
predicted_1<- ifelse(predicted_1 == 0, "laml", "lgg")
table(predicted_1) # LAML for individual 1
```

```{r}
# Load the data for the individual to predict on
individual_2 <- data.frame(unknown_data_scaled[,2])
colnames(individual_2) <- "tpm_unstranded"

input_data_2 <- matrix(0, nrow = nrow(individual_2), ncol = 6)
input_data_2[,4] <- individual_2$tpm_unstranded
rownames(input_data_2) <- rownames(individual_2)
input_data_2 <- data.frame(input_data_2)
colnames(input_data_2) <- c("unstranded", "stranded_first", "stranded_second", "tpm_unstranded",
                            "fpkm_unstranded", "fpkm_uq_unstranded")
input_matrix_2 <- as.matrix(input_data_2)

# Make predictions using the model
predictions_2 <- predict(fit, s = lambda_opt, newx = input_matrix_2, type = "response")
predicted_2 <- ifelse(predictions_2 > 0.5, 1, 0)
predicted_2 <- ifelse(predicted_2 == 0, "laml", "lgg")
table(predicted_2) # LAML for individual 2



# Load the data for the individual to predict on
individual_3<- data.frame(unknown_data_scaled[,3])
colnames(individual_3) <- "tpm_unstranded"

input_data_3 <- matrix(0, nrow = nrow(individual_3), ncol = 6)
input_data_3[,4] <- individual_3$tpm_unstranded
rownames(input_data_3) <- rownames(individual_3)
input_data_3 <- data.frame(input_data_3)
colnames(input_data_3) <- c("unstranded", "stranded_first", "stranded_second", "tpm_unstranded",
                            "fpkm_unstranded", "fpkm_uq_unstranded")
input_matrix_3 <- as.matrix(input_data_3)

# Make predictions using the model
predictions_3 <- predict(fit, s = lambda_opt, newx = input_matrix_3, type = "response")
predicted_3 <- ifelse(predictions_3 > 0.5, 1, 0)
predicted_3 <- ifelse(predicted_3 == 0, "laml", "lgg")
table(predicted_3) # LAML for individual 3



# Load the data for the individual to predict on
individual_4 <- data.frame(unknown_data_scaled[,4])
colnames(individual_4) <- "tpm_unstranded"

input_data_4 <- matrix(0, nrow = nrow(individual_4), ncol = 6)
input_data_4[,4] <- individual_4$tpm_unstranded
rownames(input_data_4) <- rownames(individual_4)
input_data_4 <- data.frame(input_data_4)
colnames(input_data_4) <- c("unstranded", "stranded_first", "stranded_second", "tpm_unstranded",
                            "fpkm_unstranded", "fpkm_uq_unstranded")
input_matrix_4 <- as.matrix(input_data_4)

# Make predictions using the model
predictions_4 <- predict(fit, s = lambda_opt, newx = input_matrix_4, type = "response")
predicted_4 <- ifelse(predictions_4 > 0.5, 1, 0)
predicted_4 <- ifelse(predicted_4 == 0, "laml", "lgg")
table(predicted_4) # LAML for individual 4



# Load the data for the individual to predict on
individual_5<- data.frame(unknown_data_scaled[,5])
colnames(individual_5) <- "tpm_unstranded"

input_data_5 <- matrix(0, nrow = nrow(individual_5), ncol = 6)
input_data_5[,4] <- individual_5$tpm_unstranded
rownames(input_data_5) <- rownames(individual_5)
input_data_5 <- data.frame(input_data_5)
colnames(input_data_5) <- c("unstranded", "stranded_first", "stranded_second", "tpm_unstranded",
                            "fpkm_unstranded", "fpkm_uq_unstranded")
input_matrix_5 <- as.matrix(input_data_5)

# Make predictions using the model
predictions_5 <- predict(fit, s = lambda_opt, newx = input_matrix_5, type = "response")
predicted_5 <- ifelse(predictions_5 > 0.5, 1, 0)
predicted_5 <- ifelse(predicted_5 == 0, "laml", "lgg")
table(predicted_5) # LAML for individual 5



# Load the data for the individual to predict on
individual_6 <- data.frame(unknown_data_scaled[,6])
colnames(individual_6) <- "tpm_unstranded"

input_data_6 <- matrix(0, nrow = nrow(individual_6), ncol = 6)
input_data_6[,4] <- individual_6$tpm_unstranded
rownames(input_data_6) <- rownames(individual_6)
input_data_6 <- data.frame(input_data_6)
colnames(input_data_6) <- c("unstranded", "stranded_first", "stranded_second", "tpm_unstranded",
                            "fpkm_unstranded", "fpkm_uq_unstranded")
input_matrix_6 <- as.matrix(input_data_6)

# Make predictions using the model
predictions_6 <- predict(fit, s = lambda_opt, newx = input_matrix_6, type = "response")
predicted_6 <- ifelse(predictions_6 > 0.5, 1, 0)
predicted_6 <- ifelse(predicted_6 == 0, "laml", "lgg")
table(predicted_6) # LAML for individual 6



# Load the data for the individual to predict on
individual_7 <- data.frame(unknown_data_scaled[,7])
colnames(individual_7) <- "tpm_unstranded"

input_data_7 <- matrix(0, nrow = nrow(individual_7), ncol = 6)
input_data_7[,4] <- individual_7$tpm_unstranded
rownames(input_data_7) <- rownames(individual_7)
input_data_7 <- data.frame(input_data_7)
colnames(input_data_7) <- c("unstranded", "stranded_first", "stranded_second", "tpm_unstranded",
                            "fpkm_unstranded", "fpkm_uq_unstranded")
input_matrix_7 <- as.matrix(input_data_7)

# Make predictions using the model
predictions_7 <- predict(fit, s = lambda_opt, newx = input_matrix_7, type = "response")
predicted_7 <- ifelse(predictions_7 > 0.5, 1, 0)
predicted_7 <- ifelse(predicted_7 == 0, "laml", "lgg")
table(predicted_7) # LAML for individual 7



# Load the data for the individual to predict on
individual_8 <- data.frame(unknown_data_scaled[,8])
colnames(individual_8) <- "tpm_unstranded"

input_data_8 <- matrix(0, nrow = nrow(individual_8), ncol = 6)
input_data_8[,4] <- individual_8$tpm_unstranded
rownames(input_data_8) <- rownames(individual_8)
input_data_8 <- data.frame(input_data_8)
colnames(input_data_8) <- c("unstranded", "stranded_first", "stranded_second", "tpm_unstranded",
                            "fpkm_unstranded", "fpkm_uq_unstranded")
input_matrix_8 <- as.matrix(input_data_8)

# Make predictions using the model
predictions_8 <- predict(fit, s = lambda_opt, newx = input_matrix_8, type = "response")
predicted_8 <- ifelse(predictions_8 > 0.5, 1, 0)
predicted_8 <- ifelse(predicted_8 == 0, "laml", "lgg")
table(predicted_8) # LAML for individual 8



# Load the data for the individual to predict on
individual_9 <- data.frame(unknown_data_scaled[,9])
colnames(individual_9) <- "tpm_unstranded"

input_data_9 <- matrix(0, nrow = nrow(individual_9), ncol = 6) 
input_data_9[,4] <- individual_9$tpm_unstranded
rownames(input_data_9) <- rownames(individual_9)
input_data_9 <- data.frame(input_data_9)
colnames(input_data_9) <- c("unstranded", "stranded_first", "stranded_second", "tpm_unstranded",
                            "fpkm_unstranded", "fpkm_uq_unstranded")
input_matrix_9 <- as.matrix(input_data_9)

# Make predictions using the model
predictions_9 <- predict(fit, s = lambda_opt, newx = input_matrix_9, type = "response")
predicted_9 <- ifelse(predictions_9 > 0.5, 1, 0)
predicted_9 <- ifelse(predicted_9 == 0, "laml", "lgg")
table(predicted_9) # LAML for individual 9

# Load the data for the individual to predict on
individual_10 <- data.frame(unknown_data_scaled[,10])
colnames(individual_10) <- "tpm_unstranded"

input_data_10 <- matrix(0, nrow = nrow(individual_10), ncol = 6)
input_data_10[,4] <- individual_10$tpm_unstranded
rownames(input_data_10) <- rownames(individual_10)
input_data_10 <- data.frame(input_data_10)
colnames(input_data_10) <- c("unstranded", "stranded_first", "stranded_second", "tpm_unstranded",
                            "fpkm_unstranded", "fpkm_uq_unstranded")
input_matrix_10 <- as.matrix(input_data_10)

# Make predictions using the model
predictions_10 <- predict(fit, s = lambda_opt, newx = input_matrix_10, type = "response")
predicted_10 <- ifelse(predictions_10 > 0.5, 1, 0)
predicted_10 <- ifelse(predicted_10 == 0, "laml", "lgg")
table(predicted_10) # LAML for individual 10
```

- Our model predicted more LAML than LGG cancer type predictions in each individual's case, meaning that LAML was a predicted cancer type for all individuals. This seems to be suspicious, even though we've obtained a high AUC value. This would suggest that data processing or merging data into one data frame for training the model had some errors. 

```{r}

classifier = randomForest(x = train_data[,1:6],
                          y = train_data$cancer_type,
                          ntree = 500, random_state = 0)

y_pred = predict(classifier, newdata = test_data[,1:6])

### making the confusion matrix
cm = table(test_data[,7], y_pred)
cm
# 0.926
# create a list to store the predicted classes for each individual
predictions <- list()

# loop through each individual from 1 to 10
for (i in 1:10) {
  # generate the input matrix for this individual
  input_matrix <- get(paste0("input_matrix_", i))
  
  # make the prediction using the trained classifier
  unknown_pred <- predict(classifier, newdata = input_matrix)
  
  # store the predicted class in the list
  predictions[[i]] <- unknown_pred
}

# create a table to display the predicted classes for all individuals
results_table <- table(data.frame(Individual = 1:10, Predicted_Class = unlist(predictions)))
results_table
```

- Again, it seems that every individual has been diagnosed with cancer type 0 which is in this case LAML, and it was the same result we obtained by performing the regression model. 

# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
